Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-09-24 move_schedules.tex
Updates  prior to moving the insertion/deletion composition into place. Included new section to clarify sorting procedure. Just need to finish revising end of Section 3 to include treaps, possibly redo pseudocode for move_left, consider whether the proof involving the constant factor should be removed or reworded, maybe add new results on bulk matrix repairing, add code links and ams classification, possibly edit the entire paper to shorten it dramatically and then it should be good to go

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition, and its complexity is often stated in this way. Informally, since column operations are the largest complexity operations needed and each column can have potentially $m$ entries, the complexity of vineyards is $O(m)$ per transposition. 
 However, there are a number of subtleties involved in this statement. 
 The first subtlety is that both the $V$ and $R$ matrices tend to be quite sparse, and thus column and row operations rarely require $m$ operations---in fact, as a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. A second subtlety, due to the necessity of sparse matrix data structures, is that querying the non-zero status of any particular entry typically takes $O(\log m)$ time. This hidden cost may seem minor, however, the quadratic scaling of the number of transpositions vineyards typically requires makes this cost non-trivial.   
In fact, achieving $O(m)$ complexity per transposition requires a special sparse matrix representation that allows swapping any two rows and columns in $O(1)$---see the appendix for more details and discussion. Finally, note the constant factor associated with the stated complexity: inspection of the individual cases of Algorithm~\ref{alg:tr} shows that there are at most two $O(m)$ operations, on both $R$ and $V$, that may be needed for any single transposition.

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[leftmargin=2\parindent, align=left, labelsep=-5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in any matrix representation occurs once per move.
 	\item[M2:] The decomposition $R = D V$ is not maintained during the movement of $\sigma_i \mapsto \sigma_j$, but only when the move is completed.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $PVP^T$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in the original vineyards publication~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times, which is excessive. If a row-oriented sparse matrix implementation is used, a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

Property M2 indicates move operations only guarantee the decomposition is valid after the operation completes, implying that the decomposition is not fully maintained during the execution of \textit{RestoreRight} and \textit{RestoreLeft} below. 
This is a significant difference compared to the vineyards algorithm, which was designed specifically to maintain a valid decomposition after each transposition. 
Thus, move operations can be thought of as making a tradeoff in granularity: whereas a sequence of adjacent transpositions $(i, i{+}1), (i{+}1, i{+}2), \dots, (j{-}1, j)$ generates $\lvert i - j \rvert $ valid decompositions in vineyards, a move operation $\mathrm{Move}(i,j)$ generates only one.

\begin{algorithm}
	\caption{Move Right Algorithm}\label{alg:mr}
	\setstretch{1.15}
    \begin{algorithmic}[1]
        \Function{RestoreRight}{$R$, $V$, $\mathbb{I} = \{I_1, I_2, \dots, I_s \}$}
            \State $(\, d_{low}, \, d_R, \, d_V \,) \gets ( \, \mathrm{low}_R(I_1), \, \mathrm{col}_R(I_1), \, \mathrm{col}_V(I_1) \, )$
        	\For{$k$ \textbf{in} $I_2, \dots, I_s$} 
        		\State $( \, d_{low}', \, d_R', \, d_V' \, ) \gets ( \, \mathrm{low}_R(k), \, \mathrm{col}_R(k), \, \mathrm{col}_V(k) \, )$
        		\State $( \, \mathrm{col}_R(k), \, \mathrm{col}_V(k) \, ) \mathrel{+}= (\, d_R, \, d_V \,) $
        		\If {$d_{low}' < d_{low}$}
        			\State $(\, d_{low}, \, d_R, \, d_V \,) \gets (\, d_{low}', \, d_R', \, d_V' \, )$
        		\EndIf
        	\EndFor 
        	\State \Return $(\, R, \, V, \, d_R, \, d_V \, )$
        \EndFunction
    \end{algorithmic}
    \begin{algorithmic}[1]
        \Function{MoveRight}{$R$, $V$, $i$, $j$} 
            \State $\mathbb{I} =$ columns satisfying $V[i, i:j] \neq 0$ %\Comment{$O(m)$}
            \State $\mathbb{J} = $ columns satisfying $\mathrm{low}_R \in [i:j]$ and $\mathrm{row}_R(i) \neq 0$ %\Comment{$O(m)$}
            \State $(\, R, \, V, d_R, d_V \,) \gets $ \Call{RestoreRight}{$R$, $V$, $\mathbb{I}$} 
            %\Comment{Restores $p$-simplices} \Comment{$O(\lvert \mathbb{I} \rvert)$}
            \State $(\, R, \, V \,) \gets $ \Call{RestoreRight}{$R$, $V$, $\mathbb{J}$}
            	%\Comment{Restores $(p+1)$-simplices}
            	%\Comment{$O(m \lvert \mathbb{I} \rvert)$}
            \State $(\, R, \, V\, ) \gets (\, P R P^T, \, P V P^T\,)$  % \Comment{$P$ is a permutation given by (i,j)}
            \State $(\, \mathrm{col}_R(j), \, \mathrm{col}_V(j) \, ) \gets (\, P d_R, \, P d_V \,)$
            \State \Return $(\, R, \, V\,)$
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\end{algorithm}

If one has a pair of filtrations $(K_0, K_1)$ each with $m$ simplices, and we assume that each simplex $\sigma_i$ switches its relative ordering with another simplex $\sigma_j$ at most once in some ordered schedule of transpositions, then the number of intermediate persistence diagrams is bounded above by $O(m^2)$ using transpositions, whereas with move operations this bound  is at most $O(m)$.  
\\
\\
\noindent \textbf{Example:} We continue the example used in sections~\ref{sec:vineyards} and~\ref{sec:reduction} to illustrate moves. Consider moving the edge $a$ to the position of edge $c$ in the filtration. The donor columns introduced in~\cite{busaryev2010tracking} are shown on the left side of each matrix. Note that using vineyards, the equivalent permutation to the decomposition would require $4$ column operations on both $R_1$ and $V_1$, respectively, whereas a single move operation accomplishes this permutation using only $2$ column operations per matrix. 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.95}
	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
	%
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		u \kern \rowsp & \; 1 \; \topstrut \\
  		v \kern \rowsp & \;  \; \\
  		w \kern \rowsp & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{R_1}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern \rowsp & \; 1 & 1 &  \; \topstrut \\
  		v \kern \rowsp & \;  & 1 &  \; \\
  		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \;  1 \; \\
  		 & \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &    &  \; \topstrut \\
  		  & \;    & 1 &  \; \\
  		 & \; 1 & 1 &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \;  \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \; 
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		  & \; 1   &    &   1  \; \topstrut \\
  		 & \;       &  1  &  1  \; \\
  		 & \; 1   &   1   &     \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} 
\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \; \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		  & \;  & 1 &  1\; \topstrut \\
  		 & \; 1 & 1 &  \; \\
  		 & \; 1 &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{d_R} 
\begin{blockarray}{cccc}{R_1}
\kern-0.2em & b & c & a  \\
	\begin{block}{c[ccc]}
		u \kern \rowsp & \;  & 1 &  \; \topstrut \\
		v \kern \rowsp & \; 1 & 1 &  \; \\
		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
	\end{block}
	% \undermat{B}{ & & }  
\end{blockarray}
\end{displaymath}

@@

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Ordered set of filtrations $\mathcal{K}$, each of size $m$, with bijections $f_{i}: K_i \to K_{i+1}$
    	\Ensure Valid $R = D V$ decompositions are computed for each filtration $K_1, K_2, \dots, K_n$
    	\Procedure{SparseSchedule}{$\mathcal{K} = ( K_1, K_2, \dots, K_n )$} 
    		\State $\mathcal{S} = \emptyset$
    		\For{$i = 1$\textbf{ to }$n-1$}
    		    \State $(p, q) \gets ([m], \, \mathrm{Im}\, f^\ast_{i})$ 
    			\State $\mathrm{lis}_q \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
    			\State $\mathcal{S} \gets \mathcal{S} \cup \mathrm{GreedySchedule}(p, q, \mathrm{lis}_q)$ \Comment{$O(d\log d)$, see section~\ref{sec:proxy_objective}}
    		\EndFor
	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial K_1$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$}
	    		\If{$i < j$}
	    			\State $(R, V) \gets$ \Call{MoveRight}{i, j}
	    		\Else
	    			\State $(R, V) \gets$ \Call{MoveLeft}{i, j}
	    		\EndIf
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, 
and is meant to serve as a guidepost before discussing its  components in depth. 
Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. 
Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly. Due to the combinatorial aspect of our approach, there is actually no need to have access to---or explicitly store---the entire family of filtrations and the schedules between them. That is, Algorithm~\ref{alg:schedule} can be  modified to be completely online, keeping at most two filtrations and one decomposition in memory at any given time.

@@

\subsubsection{The Continuous Setting}\label{sec:continuous_setting}
In the vineyards case, we are given a homotopy $F : K \times [0,1] \to \mathbb{R}$ and a pair of filtrations $K_0, K_1$ of $K$, with simplices  ordered according to $f_0 = F(\cdot, 0)$ and $f_1 = F(\cdot, 1)$, respectively. 
The choice of homotopy $F$ completely determines the number of adjacent transpositions, so to get a meaningful bound we must know something about $F$. If we assume that the curves $\tau \mapsto F(\cdot, \tau)$  are in general position in $[0,1]\times \mathbb{R}$, and that each pair of curves cross  at most once, then the set of adjacent transpositions $S_F$ that $F$ decomposes into is given by the pairs of simplices with changes in their relative ordering:
\begin{align}\label{eq:sf_schedule}
	S_F &= \{ (i,j) \, | \, f_0(\sigma_i) < f_0(\sigma_j) \mbox{ and }  f_1(\sigma_i) > f_1(\sigma_j) \quad  \; i,j \in [m] \} \\
	&= \{ s_1, s_2, \dots, s_k \}
\end{align}
%We assume, without loss of generality, that the underlying complexes $K_0$ and $K_1$ have the same cardinality. 
Since we deal exclusively with essential filtrations, we may think of $K_1$ as simply a reordering of $K_0$ (or vice versa). Let $p = [m]$ and let $q = \mathrm{Im}\,f^\ast$, where $f^\ast : [m] \to [m]$ is the one-to-one correspondence induced by the bijection $f: K_0 \to K_1$. Then $S_F$ is given by the inversions between $p$ and $q$:
\[
\mathrm{Inv}(p, q) = \{ \, (i,j) \, | \, p(i) < p(j) \mbox{ and } q(i) > q(j)  \quad \; i,j \in [m]\, \}
\]
The cardinality of $  S_F $ is exactly the \emph{Kendall}-$\tau$ distance $K_\tau(p,q)$ between $p$ and $q$: 
\begin{equation}\label{eq:kendall_dist}
	K_\tau(p, q) = \lvert \, \mathrm{Inv}(p, q) \, \rvert
\end{equation}
Geometrically, our assumptions on $F$ defines a class of $x$-monotone curves called \emph{pseudo-segments}, i.e. parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant. This family includes 
the straight-line homotopy $F(\sigma, \tau) = (1 - \tau) f_0(\sigma) + \tau f_1(\sigma)$,   studied in the original vineyards paper~\cite{cohen2006vines}. Detecting all $k$ intersections of $m$ pseudo-segments is a well-studied problem in computational geometry that can be optimally solved in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term. 
Since $k \sim O(m^2)$ in the worst case, achieved when $f_1 = - f_0$, we have the following upper bound on $\lvert S_F \rvert$
\begin{equation}\label{eq:sf_m2}
	\lvert S_F \rvert \approx O(m^2)
\end{equation}   
 This bound is exhibited in the grayscale image data example in section~\ref{sec:motivation}: each 9x9 patch contains $(81, 208, 128)$ simplices of dimensions $(0, 1, 2)$ yielding $m=417$ simplices in total which, as the homotopy is simulated, generate $\approx 70,\!000$ transpositions. The upper bound size on $\lvert S_F \rvert$ is thus $\binom{417}{2} \approx 87,\!000$, which nearly occurs in this case, since the last filtration is almost the reverse of the filtration at the beginning.   
 This quadratic scaling can induce a number of issues in practical implementations of vineyards.
 Indeed, although many transpositions require $0$ column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time. Since this check is required after each permutations are applied, it is in some sense unavoidable in the vineyards algorithm.

>>>>

%Prior to performing the exchange, observe that that the bold entry in $V_1$ would render $V_1$ non-upper triangular after the exchange. This entry is removed by a left-to-right column operation, given by applying the $S_1^{2}$ on the right to $R_1$ and $V_1$. After this operation, the permutation may be safely applied to $V_1$. In this case, both before and after the permutation, $R_1$ is rendered non-reduced, requiring another column operation to restore the decomposition to a valid state.

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition, and its complexity is often stated in this way. Informally, since column operations are the largest complexity operations needed and each column can have potentially $m$ entries, the complexity of vineyards is $O(m)$ per transposition. 
 However, there are a number of subtleties involved in this statement. 
 The first subtlety is that both the $V$ and $R$ matrices tend to be quite sparse, and thus column and row operations rarely require $m$ operations---in fact, as a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. A second subtlety, due to the necessity of sparse matrix data structures, is that querying the non-zero status of any particular entry typically takes $O(\log m)$ time. This hidden cost may seem minor, however, the quadratic scaling of the number of transpositions vineyards typically requires makes this cost non-trivial.   
In fact, achieving $O(m)$ complexity per transposition requires a special sparse matrix representation that allows swapping any two rows and columns in $O(1)$---see the appendix for more details and discussion. Finally, note the constant factor associated with the stated complexity: inspection of the individual cases of the algorihtm from~\cite{cohen2006vines} shows that there are at most two $O(m)$ operations, on both $R$ and $V$, that may be needed for any single transposition.

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[leftmargin=2\parindent, align=left, labelsep=-5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in either $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is not maintained during the movement of $\sigma_i \mapsto \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in the original vineyards publication~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times, which is excessive. If a row-oriented sparse matrix implementation is used, a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

Property M2 indicates move operations only guarantee the decomposition is valid after the operation completes, implying that the decomposition is not fully maintained during the execution of \textit{RestoreRight} and \textit{RestoreLeft} below. 
This is a significant difference compared to the vineyards algorithm, which was designed specifically to maintain a valid decomposition after each transposition. 
In this way, we interpret move operations as making a tradeoff in granularity: whereas a sequence of adjacent transpositions $(i, i{+}1), (i{+}1, i{+}2), \dots, (j{-}1, j)$ generates $\lvert i - j \rvert $ valid decompositions in vineyards, a move operation $\mathrm{Move}(i,j)$ generates only one.
Indeed, if one has a pair of filtrations $(K_0, K_1)$ each with $m$ simplices, and we assume that each simplex $\sigma_i$ switches its relative ordering with another simplex $\sigma_j$ at most once in some ordered schedule of transpositions, then the number of intermediate persistence diagrams is bounded above by $O(m^2)$ using transpositions, whereas with move operations this bound clearly is at most $O(m)$.  
\begin{algorithm}
	\caption{Move Right Algorithm}\label{alg:mr}
	\setstretch{1.15}
    \begin{algorithmic}[1]
        \Function{RestoreRight}{$R$, $V$, $\mathbb{I} = \{I_1, I_2, \dots, I_s \}$}
            \State $(\, d_{low}, \, d_R, \, d_V \,) \gets ( \, \mathrm{low}_R(I_1), \, \mathrm{col}_R(I_1), \, \mathrm{col}_V(I_1) \, )$
        	\For{$k$ \textbf{in} $I_2, \dots, I_s$} 
        		\State $( \, d_{low}', \, d_R', \, d_V' \, ) \gets ( \, \mathrm{low}_R(k), \, \mathrm{col}_R(k), \, \mathrm{col}_V(k) \, )$
        		\State $( \, \mathrm{col}_R(k), \, \mathrm{col}_V(k) \, ) \mathrel{+}= (\, d_R, \, d_V \,) $
        		\If {$d_{low}' < d_{low}$}
        			\State $(\, d_{low}, \, d_R, \, d_V \,) \gets (\, d_{low}', \, d_R', \, d_V' \, )$
        		\EndIf
        	\EndFor 
        	\State \Return $(\, R, \, V, \, d_R, \, d_V \, )$
        \EndFunction
    \end{algorithmic}
    \begin{algorithmic}[1]
        \Function{MoveRight}{$R$, $V$, $i$, $j$} 
            \State $\mathbb{I} =$ columns satisfying $V[i, i:j] \neq 0$ %\Comment{$O(m)$}
            \State $\mathbb{J} = $ columns satisfying $\mathrm{low}_R \in [i:j]$ and $\mathrm{row}_R(i) \neq 0$ %\Comment{$O(m)$}
            \State $(\, R, \, V, d_R, d_V \,) \gets $ \Call{RestoreRight}{$R$, $V$, $\mathbb{I}$} 
            %\Comment{Restores $p$-simplices} \Comment{$O(\lvert \mathbb{I} \rvert)$}
            \State $(\, R, \, V \,) \gets $ \Call{RestoreRight}{$R$, $V$, $\mathbb{J}$}
            	%\Comment{Restores $(p+1)$-simplices}
            	%\Comment{$O(m \lvert \mathbb{I} \rvert)$}
            \State $(\, R, \, V\, ) \gets (\, P R P^T, \, P V P^T\,)$  % \Comment{$P$ is a permutation given by (i,j)}
            \State $(\, \mathrm{col}_R(j), \, \mathrm{col}_V(j) \, ) \gets (\, P d_R, \, P d_V \,)$
            \State \Return $(\, R, \, V\,)$
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\end{algorithm}
If one has a pair of filtrations $(K_0, K_1)$ each with $m$ simplices, and we assume that each simplex $\sigma_i$ switches its relative ordering with another simplex $\sigma_j$ at most once in some ordered schedule of transpositions, then the number of intermediate persistence diagrams is bounded above by $O(m^2)$ using transpositions, whereas with move operations this bound  is at most $O(m)$.  
\\
\\
\noindent \textbf{Example:} We continue the example used in sections~\ref{sec:vineyards} and~\ref{sec:reduction} to illustrate moves. Consider moving the edge $a$ to the position of edge $c$ in the filtration. The donor columns introduced in~\cite{busaryev2010tracking} are shown on the left side of each matrix. Note that using vineyards, the equivalent permutation to the decomposition would require $4$ column operations on both $R_1$ and $V_1$, respectively, whereas a single move operation accomplishes this permutation using only $2$ column operations per matrix. 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.95}
	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
	%
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		u \kern \rowsp & \; 1 \; \topstrut \\
  		v \kern \rowsp & \;  \; \\
  		w \kern \rowsp & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{R_1}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern \rowsp & \; 1 & 1 &  \; \topstrut \\
  		v \kern \rowsp & \;  & 1 &  \; \\
  		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \;  1 \; \\
  		 & \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &    &  \; \topstrut \\
  		  & \;    & 1 &  \; \\
  		 & \; 1 & 1 &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \;  \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \; 
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		  & \; 1   &    &   1  \; \topstrut \\
  		 & \;       &  1  &  1  \; \\
  		 & \; 1   &   1   &     \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} 
\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \; \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		  & \;  & 1 &  1\; \topstrut \\
  		 & \; 1 & 1 &  \; \\
  		 & \; 1 &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{d_R} 
\begin{blockarray}{cccc}{R_1}
\kern-0.2em & b & c & a  \\
	\begin{block}{c[ccc]}
		u \kern \rowsp & \;  & 1 &  \; \topstrut \\
		v \kern \rowsp & \; 1 & 1 &  \; \\
		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
	\end{block}
	% \undermat{B}{ & & }  
\end{blockarray}
\end{displaymath}

@@

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Ordered set of filtrations $\mathcal{K}$, each of size $m$, with bijections $f_{i}: K_i \to K_{i+1}$
    	\Ensure Valid $R = D V$ decompositions are computed for each filtration $K_1, K_2, \dots, K_n$
    	\Procedure{CoarseSchedule}{$\mathcal{K} = ( K_1, K_2, \dots, K_n )$} 
    		\State $\mathcal{S} = \emptyset$
    		\For{$i = 1$\textbf{ to }$n-1$}
    		    \State $(p, q) \gets ([m], \, \mathrm{Im}\, f^\ast_{i})$ 
    			\State $\mathrm{lis}_q \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
    			\State $\mathcal{S} \gets \mathcal{S} \cup \mathrm{GreedySchedule}(p, q, \mathrm{lis}_q)$ \Comment{$O(d\log m)$, see section~\ref{sec:proxy_objective}}
    		\EndFor
	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial K_1$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$}
	    		\If{$i < j$}
	    			\State $(R, V) \gets$ \Call{MoveRight}{i, j}
	    		\Else
	    			\State $(R, V) \gets$ \Call{MoveLeft}{i, j}
	    		\EndIf
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, 
and is meant to serve as a guidepost before discussing its  components in depth. 
Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. 
Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly. Due to the combinatorial aspect of our approach, there is actually no need to have access to---or explicitly store---the entire family of filtrations and the schedules between them. That is, Algorithm~\ref{alg:schedule} can be  modified to be completely online, keeping at most two filtrations and one decomposition in memory at any given time.

@@

\subsubsection{The Continuous Setting}\label{sec:continuous_setting}
In the vineyards case, we are given a homotopy $F : K \times [0,1] \to \mathbb{R}$ and a pair of filtrations $K_0, K_1$ of $K$, with simplices  ordered according to $f_0 = F(\cdot, 0)$ and $f_1 = F(\cdot, 1)$, respectively. 
The choice of homotopy $F$ completely determines the number of adjacent transpositions, so to get a meaningful bound we must know something about $F$. If we assume that the curves $\tau \mapsto F(\cdot, \tau)$  are in general position in $[0,1]\times \mathbb{R}$, and that each pair of curves cross  at most once, then the set of adjacent transpositions $S_F$ that $F$ decomposes into is given by the pairs of simplices with changes in their relative ordering:
\begin{align}\label{eq:sf_schedule}
	S_F &= \{ (i,j) \, | \, f_0(\sigma_i) < f_0(\sigma_j) \mbox{ and }  f_1(\sigma_i) > f_1(\sigma_j) \quad  \; i,j \in [m] \} \\
	&= \{ s_1, s_2, \dots, s_k \}
\end{align}
%We assume, without loss of generality, that the underlying complexes $K_0$ and $K_1$ have the same cardinality. 
Since we deal exclusively with essential filtrations, we may think of $K_1$ as simply a reordering of $K_0$ (or vice versa). Let $p = [m]$ and let $q = \mathrm{Im}\,f^\ast$, where $f^\ast : [m] \to [m]$ is the one-to-one correspondence induced by the bijection $f: K_0 \to K_1$. Then $S_F$ is given by the inversions between $p$ and $q$:
\[
\mathrm{Inv}(p, q) = \{ \, (i,j) \, | \, p(i) < p(j) \mbox{ and } q(i) > q(j)  \quad \; i,j \in [m]\, \}
\]
The cardinality of $  S_F $ is exactly the \emph{Kendall}-$\tau$ distance $K_\tau(p,q)$ between $p$ and $q$: 
\begin{equation}\label{eq:kendall_dist}
	K_\tau(p, q) = \lvert \, \mathrm{Inv}(p, q) \, \rvert
\end{equation}
Geometrically, our assumptions on $F$ defines a class of $x$-monotone curves called \emph{pseudo-segments}, i.e. parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant. This family includes 
the straight-line homotopy $F(\sigma, \tau) = (1 - \tau) f_0(\sigma) + \tau f_1(\sigma)$,   studied in the original vineyards paper~\cite{cohen2006vines}. Detecting all $k$ intersections of $m$ pseudo-segments is a well-studied problem in computational geometry that can be optimally solved in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term. 
Since $k \sim O(m^2)$ in the worst case, achieved when $f_1 = - f_0$, we have the following upper bound on $\lvert S_F \rvert$
\begin{equation}\label{eq:sf_m2}
	\lvert S_F \rvert \approx O(m^2)
\end{equation}   
 This bound is exhibited in the grayscale image data example in section~\ref{sec:motivation}: each 9x9 patch contains $(81, 208, 128)$ simplices of dimensions $(0, 1, 2)$ yielding $m=417$ simplices in total which, as the homotopy is simulated, generate $\approx 70,\!000$ transpositions. The upper bound size on $\lvert S_F \rvert$ is thus $\binom{417}{2} \approx 87,\!000$, which nearly occurs in this case, since the last filtration is approximately the reverse of the filtration at the beginning.   
 This quadratic scaling can induce a number of issues in practical implementations of vineyards.
 Indeed, although many transpositions require $0$ column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time. Since this check is required after each permutations are applied, it is in some sense unavoidable in the vineyards algorithm.

The ordered set of inversions $(i,j) \in S_F$ can be interpreted as permutations to apply to the decomposition. We call $S_F$ a \emph{schedule} with respect to $F$. If our goal is to decrease the size of $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by collapsing contiguous sequences of adjacent transpositions to moves, via the map:
\begin{equation}\label{eq:tr_to_mv}
	 (i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)  \quad \text{ if } i < j
\end{equation}
We expect $\widetilde{S}_F$ to be cheaper to execute than $S_F$, with the tradeoff  that less diagrams are produced.
We make this precise next:
\begin{proposition}\label{prop:factor2}
	Let $R = D V$ be a valid decomposition of size $(m \times m)$, let $C$ denote the number of $O(m)$ operations required to execute schedule $S$, and similarly let $L$ be the number of $O(m)$ operations required to execute a coarsened schedule $\widetilde{S}$ of $S$. We have the following relationship between $C$ and $L$:
	$$ \frac{C}{2} \leq L \leq C$$ 
\end{proposition}
\begin{proof}
Consider executing the vineyards algorithm on a given schedule $S$.
	%Consider Algorithms~\ref{alg:tr} and~\ref{alg:mr}, where Algorithm~\ref{alg:tr} is used to execute a given schedule $S$ and Algorithm~\ref{alg:mr} is used to execute its coarsened version $\widetilde{S}$. If maps are created providing $O(1)$ access to the low entries needed by lines (5) and (12), and the non-zero entries in each column are sorted according to some fixed order providing $O(\log(m))$ time for lines (5), (11), (17), and (20), then  
	%If we maintain a map locating the position of the lowest entries in $R$ in $O(1)$ time, 
	The dominant subcomputation for the vineyards algorithm are the column operations---which take $O(m)$ time---since the permutations can be made to the matrix in $O(1)$ time and all other computations take at most $O(\log m)$ time. In the vineyards algorithm, there are at most $2$ column operations for a given adjacent transposition per matrix. 
	Thus, for any contiguous sequence of transpositions $(t_i, t_{i{+}1}, \dots, t_j)$ we require at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.
